---
title: "Alcohol Consumption and Average Grades"
author: "Karan Vijay Kashyap, Sandile Nhlalo-Sibanda"
date: "August, 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\centering

\raggedright

\abstract
This project investigates if alcohol intake has any predictive value over average student grades. We're also curious about replicating the original study and seeing whether other factors may be major predictors of student grades. We will not attempt to forecast student grade progression across marking periods because all of the characteristics in the dataset (except grades) remain constant throughout marking periods and are basic descriptors of student backgrounds.

\newpage 
# Introduction
The project is set up as follows.
We first do Exploratory Data Analysis to determine whether the two tables (with math and Portuguese grades) can be integrated.

We derive student average grades throughout grading periods by integrating the two tables and estimate three models: Linear Model, Regression Tree Model, and Random Forest Model.

Later, we acquired data from a survey for English and Math grades, and we will replicate the research on a fresh dataset, estimating three models: Linear Model, Regression Tree Model, and Random Forest Model.

# About Data
While this is an excellent dataset, it lacks information on parents' income. These findings are based on surveys of students done at two Portuguese schools. As a result, I assume there shouldn't be much variety in parents' wealth (definitely not as much as in the California education system), class sizes, and computers/students. As a result, one downside (or advantage, depending on your point of view) of this dataset is that it maintains some of the crucial variables relatively constant.
Schoolsup is a characteristic in the dataset that may be associated to parental income (extra educational support, yes or no). After higher and Medu, schoolsup is the third most important predictor of student average grades. What's more, schoolsup is inversely associated to average grades.


## Objective of study
The goal of this study is to determine the effect of alcohol use on students' average grades.

\newpage

# Load the packages

```{r, include=FALSE}
library(ggplot2)
library(plyr)
library(dplyr)
library(gridExtra)
library(alluvial)
library(extrafont)
library(readr)
library(rpart)
library(waffle)
library(party)
library(randomForest)
```


# Data Analysis

```{r, echo=FALSE}
d1 <- read.table("student-mat.csv",sep=",",header=TRUE) #Reading Math dataset
d2 <- read.table("student-por.csv",sep=",",header=TRUE) #Reading Portuguese dataset

#****New survey data
data_english <- read.table("Student_English.csv",sep=",",header=TRUE) #Reading English dataset
data_maths <- read.table("Student_Maths.csv",sep=",",header=TRUE) #Reading Maths dataset

colnames(data_maths)[1] <- "school"
colnames(data_english)[1] <- "school"
```


```{r, echo=FALSE}
#merging datasets by common columns into single DS
data.source=merge(d1,d2,by=c("school","sex","age","address", "famsize", "Pstatus", "Medu", "Fedu", "Mjob", "Fjob", "reason", "nursery", "internet", "guardian","guardian","traveltime","studytime","failures", "schoolsup","famsup","activities","higher","romantic", "famrel", "freetime", "goout", "Dalc", "Walc", "health", "absences"))

head(data.source)
```

```{r, echo=FALSE}
#merging NEW datasets by common columns into single DS
new_data.source=merge(data_english,data_maths,by=c("school","sex","age","address","famsize","Pstatus", "Medu", "Fedu", "Mjob", "Fjob", "reason", "nursery", "internet", "guardian", "guardian", "traveltime", "studytime", "failures", "schoolsup", "famsup", "activities", "higher", "romantic", "famrel", "freetime", "goout", "Dalc", "Walc", "health", "absences"))

head(new_data.source)
```

## Including Plots

####You can also embed plots, for example:

```{r, echo=FALSE}
#Research Data + New Survey data
data.source$mathgrades=rowMeans(cbind(data.source$G1.x,data.source$G2.x,data.source$G3.x))
data.source$portgrades=rowMeans(cbind(data.source$G1.y,data.source$G2.y,data.source$G3.y))

data.source$Dalc <- as.factor(data.source$Dalc)      
data.source$Dalc <- mapvalues(data.source$Dalc, 
                              from = 1:5, 
                              to = c("Very Low", "Low", "Medium", "High", "Very High"))

str1=ggplot(data.source, aes(x=mathgrades, y=portgrades)) +
 geom_point(aes(colour=factor(Dalc)))+ scale_colour_hue(l=25,c=150)+
geom_smooth(method = "lm", se = FALSE)

data.source$Walc <- as.factor(data.source$Walc)      
data.source$Walc <- mapvalues(data.source$Walc, 
                              from = 1:5, 
                              to = c("Very Low", "Low", "Medium", "High", "Very High"))

str2=ggplot(data.source, aes(x=mathgrades, y=portgrades))+
geom_point(aes(colour=factor(Walc)))+ scale_colour_hue(l=25,c=150)+
geom_smooth(method = "lm", se = FALSE)
########################################################################
new_data.source$mathgrades=rowMeans(cbind(new_data.source$G1.x,new_data.source$G2.x,new_data.source$G3.x))
new_data.source$enggrades=rowMeans(cbind(new_data.source$G1.y,new_data.source$G2.y,new_data.source$G3.y))

new_data.source$Dalc <- as.factor(new_data.source$Dalc)      
new_data.source$Dalc <- mapvalues(new_data.source$Dalc, 
                              from = 1:5, 
                              to = c("Very Low", "Low", "Medium", "High", "Very High"))

new_str1=ggplot(new_data.source, aes(x=mathgrades, y=enggrades)) +
 geom_point(aes(colour=factor(Dalc)))+ scale_colour_hue(l=25,c=150)+
geom_smooth(method = "lm", se = FALSE)

new_data.source$Walc <- as.factor(new_data.source$Walc)      
new_data.source$Walc <- mapvalues(new_data.source$Walc, 
                              from = 1:5, 
                              to = c("Very Low", "Low", "Medium", "High", "Very High"))

new_str2=ggplot(new_data.source, aes(x=mathgrades, y=enggrades))+
geom_point(aes(colour=factor(Walc)))+ scale_colour_hue(l=25,c=150)+
geom_smooth(method = "lm", se = FALSE)

grid.arrange(str1, new_str1, str2, new_str2,nrow=2)
```


```{r, echo=FALSE}
d3<-rbind(d1,d2) #combine the two datasets
# and eliminate the repeats:
d3norepeats<-d3 %>% distinct(school,sex,age,address,famsize,Pstatus,
                Medu,Fedu,Mjob,Fjob,reason,
                guardian,traveltime,studytime,failures,
                schoolsup, famsup,activities,nursery,higher,internet,
                romantic,famrel,freetime,goout,Dalc,Walc,health,absences, .keep_all = TRUE)
#add a column with average grades (math or Portuguese, whichever is available)
d3norepeats$avggrades=rowMeans(cbind(d3norepeats$G1,d3norepeats$G2,d3norepeats$G3))
# and drop grades in 3 marking periods.
d3norepeats<-d3norepeats[,-(31:33)]
```


```{r, echo=FALSE}
#New Survey Data
data_combine<-rbind(data_english,data_maths) #combine the two datasets
# and eliminate the repeats:
data_combine_norepeats<-data_combine %>% distinct(school,sex,age,address,famsize,Pstatus,
                Medu,Fedu,Mjob,Fjob,reason,
                guardian,traveltime,studytime,failures,
                schoolsup, famsup,activities,nursery,higher,internet,
                romantic,famrel,freetime,goout,Dalc,Walc,health,absences, .keep_all = TRUE)
#add a column with average grades (math or Portuguese, whichever is available)
data_combine_norepeats$avggrades=rowMeans(cbind(data_combine_norepeats$G1,data_combine_norepeats$G2,data_combine_norepeats$G3))
# and drop grades in 3 marking periods.
data_combine_norepeats<-data_combine_norepeats[,-(31:33)]
```


```{r, echo=FALSE}
ggplot(d3norepeats, aes(x=Dalc, y=avggrades, group=Dalc))+
  geom_boxplot()+
  theme(legend.position="none")+
  #scale_fill_manual(values=waffle.col)+
  xlab("Daily Alcohol consumption")+
  ylab("Average Grades")+
  ggtitle("Average Grade (Original Study)")
```


```{r, echo=FALSE}
#New Survey Data
ggplot(data_combine_norepeats, aes(x=Dalc, y=avggrades, group=Dalc))+
  geom_boxplot()+
  theme(legend.position="none")+
  #scale_fill_manual(values=waffle.col)+
  xlab("Daily Alcohol consumption")+
  ylab("Average Grades")+
  ggtitle("Average Grade (Reproducible Survey)")

```


```{r, echo=FALSE}
failureind<-which(names(d3norepeats)=="failures")
d3norepeats<-d3norepeats[,-failureind]
```
```{r, echo=FALSE}
# 1) multiple regression
lm2<-lm(avggrades~., data=d3norepeats[,1:30])
summary(lm2)
```


```{r, echo=FALSE}
#predictions
rt2<-rpart(avggrades~., data=d3norepeats[,1:30])
lm.predictions<-predict(lm2,d3norepeats)
rt.predictions<-predict(rt2,d3norepeats)
nmse.lm<-mean((lm.predictions-d3norepeats[,"avggrades"])^2)/mean((mean(d3norepeats$avggrades)-d3norepeats[,"avggrades"])^2)
nmse.rt<-mean((rt.predictions-d3norepeats[,"avggrades"])^2)/mean((mean(d3norepeats$avggrades)-d3norepeats[,"avggrades"])^2)
print(nmse.lm) #0.79
print(nmse.rt) #0.85
```

```{r, echo=FALSE}
#New Survey Data
failureind<-which(names(data_combine_norepeats)=="failures")
data_combine_norepeats<-data_combine_norepeats[,-failureind]
```
```{r, echo=FALSE}
# 1) multiple regression
lm2<-lm(avggrades~., data=data_combine_norepeats[,1:30])
summary(lm2)
```

```{r, echo=FALSE}
#New Survey Data
#predictions
rt2<-rpart(avggrades~., data=data_combine_norepeats[,1:30])
lm.predictions<-predict(lm2,data_combine_norepeats)
rt.predictions<-predict(rt2,data_combine_norepeats)
nmse.lm<-mean((lm.predictions-data_combine_norepeats[,"avggrades"])^2)/mean((mean(data_combine_norepeats$avggrades)-data_combine_norepeats[,"avggrades"])^2)
nmse.rt<-mean((rt.predictions-data_combine_norepeats[,"avggrades"])^2)/mean((mean(data_combine_norepeats$avggrades)-data_combine_norepeats[,"avggrades"])^2)
print(nmse.lm) #0.79
print(nmse.rt) #0.85
```


```{r, echo=FALSE}
lmpltdata1=data.frame(cbind(lm.predictions,d3norepeats[,"avggrades"]))
colnames(lmpltdata1)<-c("lm.predictions","avggrades")
rtpltdata1=data.frame(cbind(rt.predictions,d3norepeats[,"avggrades"]))
colnames(rtpltdata1)<-c("rt.predictions","avggrades")

d3norepeats$Dalc<-as.factor(d3norepeats$Dalc)

errplt.lt1=ggplot(lmpltdata1,aes(lm.predictions,avggrades))+
                  geom_point(aes(color=d3norepeats[,"Dalc"]))+
                  xlab("Predicted Grades (Linear Model)")+
                  ylab("Actual Grades")+
                  geom_abline(intercept=0,slope=1,color="#0066CC",size=1)+
                  #geom_smooth(method = "lm", se = FALSE)+
                  scale_colour_brewer(palette = "Set1",name = "Daily Alcohol \nConsumption")

errplt.rt1=ggplot(rtpltdata1,aes(rt.predictions,avggrades))+
  geom_point(aes(color=d3norepeats[,"Dalc"]))+
  xlab("Predicted Grades (Regression Tree)")+
  ylab("Actual Grades")+
  geom_abline(intercept=0,slope=1,color="#0066CC",size=1)+
  #geom_smooth(method = "lm", se = FALSE)+
  scale_colour_brewer(palette = "Set1",name = "Daily Alcohol \nConsumption")

grid.arrange(errplt.lt1,errplt.rt1,nrow=2)
```
```{r, echo=FALSE}
library(randomForest)
set.seed(4543)
rf2<-randomForest(avggrades~., data=d3norepeats[,1:30], ntree=500, importance=T)
rf.predictions<-predict(rf2,d3norepeats)
nmse.rf<-mean((rf.predictions-d3norepeats[,"avggrades"])^2)/mean((mean(d3norepeats$avggrades)-d3norepeats[,"avggrades"])^2)
print(nmse.rf) #0.2
```


```{r, echo=FALSE}
#New Survey Data
lmpltdata1=data.frame(cbind(lm.predictions,data_combine_norepeats[,"avggrades"]))
colnames(lmpltdata1)<-c("lm.predictions","avggrades")
rtpltdata1=data.frame(cbind(rt.predictions,data_combine_norepeats[,"avggrades"]))
colnames(rtpltdata1)<-c("rt.predictions","avggrades")

data_combine_norepeats$Dalc<-as.factor(data_combine_norepeats$Dalc)

errplt.lt1=ggplot(lmpltdata1,aes(lm.predictions,avggrades))+
                  geom_point(aes(color=data_combine_norepeats[,"Dalc"]))+
                  xlab("Predicted Grades (Linear Model)")+
                  ylab("Actual Grades")+
                  geom_abline(intercept=0,slope=1,color="#0066CC",size=1)+
                  #geom_smooth(method = "lm", se = FALSE)+
                  scale_colour_brewer(palette = "Set1",name = "Daily Alcohol \nConsumption")

errplt.rt1=ggplot(rtpltdata1,aes(rt.predictions,avggrades))+
  geom_point(aes(color=data_combine_norepeats[,"Dalc"]))+
  xlab("Predicted Grades (Regression Tree)")+
  ylab("Actual Grades")+
  geom_abline(intercept=0,slope=1,color="#0066CC",size=1)+
  #geom_smooth(method = "lm", se = FALSE)+
  scale_colour_brewer(palette = "Set1",name = "Daily Alcohol \nConsumption")

grid.arrange(errplt.lt1,errplt.rt1,nrow=2)
```
```{r, echo=FALSE}
library(randomForest)
set.seed(4543)
randfor<-randomForest(avggrades~., data=data_combine_norepeats[,1:30], ntree=500, importance=T)
rf.predictions<-predict(randfor,data_combine_norepeats)
nmse.rf<-mean((rf.predictions-data_combine_norepeats[,"avggrades"])^2)/mean((mean(data_combine_norepeats$avggrades)-data_combine_norepeats[,"avggrades"])^2)
print(nmse.rf) #0.2
```



```{r, echo=FALSE}
#first combine the rf predictions and actual scores in a single data frame
rfpltdata1=data.frame(cbind(rf.predictions,d3norepeats[,"avggrades"]))
colnames(rfpltdata1)<-c("rf.predictions","avggrades")

# then create the error plot.
errplt.rf1<-ggplot(rfpltdata1,aes(rf.predictions,avggrades))+
  geom_point(aes(color=d3norepeats[,"Dalc"]))+
  xlab("Predicted Grades (Random Forest with 500 Trees)")+
  ylab("Actual Grades")+
  geom_abline(intercept=0,slope=1,color="#0066CC",size=1)+
  #geom_smooth(method = "lm", se = FALSE)+
  scale_colour_brewer(palette = "Set1",name = "Daily Alcohol \nConsumption")
#finally, plot the error plot from the random forest with the error plots of the linear and regression tree models.
grid.arrange(errplt.rf1, errplt.lt1,errplt.rt1,nrow=3)
```
```{r, echo=FALSE}
varImpPlot(rf2,type=1) 
```



```{r, echo=FALSE}
#New Survey Data
#first combine the rf predictions and actual scores in a single data frame
rfpltdata1=data.frame(cbind(rf.predictions,data_combine_norepeats[,"avggrades"]))
colnames(rfpltdata1)<-c("rf.predictions","avggrades")

# then create the error plot.
errplt.rf1<-ggplot(rfpltdata1,aes(rf.predictions,avggrades))+
  geom_point(aes(color=data_combine_norepeats[,"Dalc"]))+
  xlab("Predicted Grades (Random Forest with 500 Trees)")+
  ylab("Actual Grades")+
  geom_abline(intercept=0,slope=1,color="#0066CC",size=1)+
  #geom_smooth(method = "lm", se = FALSE)+
  scale_colour_brewer(palette = "Set1",name = "Daily Alcohol \nConsumption")
#finally, plot the error plot from the random forest with the error plots of the linear and regression tree models.
grid.arrange(errplt.rf1, errplt.lt1,errplt.rt1,nrow=3)
```
```{r, echo=FALSE}
varImpPlot(randfor,type=1) 
```
########################################################################
########################################################################
########################################################################

